---
layout: post
title: "Profit and Loss Analysis"
date: 2016-10-15
output:
  html_document:
    css: style.css
share: true
categories: blog
excerpt: "The Usage of the #FundEIU hashtag on Twitter"
tags: [rstats]
---




```{r message=FALSE, warning=FALSE, echo=FALSE}
library(ggplot2)
library(plotly)
library(DT)
library(dplyr)

Sys.setenv("plotly_username"="ryanburge")
Sys.setenv("plotly_api_key"="bvzv16do7x")

all <- read.csv("D:/profitloss/all.csv", stringsAsFactors = FALSE, strip.white = TRUE)


all$Year <- trimws(all$Year)
all$Scholarships <- trimws(all$Scholarships)
all$Amount <- trimws(all$Amount)
all$Summer <- trimws(all$Summer)
all$Amount <- gsub("^[(]", "-", all$Amount)
all$Amount <- gsub("[)]", "", all$Amount)
all$Amount <- as.numeric(all$Amount)
all$Scholarships <- as.numeric(all$Scholarships)
all$Summer <- as.numeric(all$Summer)

all$Year <- factor(all$Year, levels=unique(all$Year))

all$deducts <- all$Waivers + all$Scholarships
all$deductper <- all$deducts/all$Total

all$Total.Expenses <- all$Net.Tuition - all$Amount
all$Other.Expenses <- all$Total.Expenses - all$Personnel
  

agg <- aggregate(all$Personnel, list(all$Year), sum)
agg2 <- aggregate(all$Net.Tuition, list(all$Year), sum)
agg3 <- aggregate(all$deducts, list(all$Year), sum)
agg4 <- aggregate(all$Total.Expenses, list(all$Year), sum)
agg$type <- c("Personnel Expenses")
agg2$type <- c("Net Tuition")
agg3$type <- c("Waivers + Scholarships")
agg4$type <- c("Total Expenses")
agg5 <- aggregate(all$Amount, list(all$Dept), sum)



total <- rbind(agg, agg2, agg3, agg4)

cah <- filter(all, College == "CAH")
cos <- filter(all, College == "COS")
bus <- filter(all, College == "Business")
ceps <- filter(all, College == "CEPS")

last <- filter(all, Year == "2015-2016")
first <- filter(all, Year == "2011-2012")
agg5$College <- last$College


last$tuition10 <- last$Total.Expenses - (last$Net.Tuition*1.1)
last$tuition10 <- last$tuition10*(-1)

last$tuition15 <- last$Total.Expenses - (last$Net.Tuition*1.15)
last$tuition15 <- last$tuition15*(-1)


displaylast <- select(last, Dept, College, Amount, Personnel, Net.Tuition, deductper)



```

###Trends in Profit and Loss by College and Department

The following graphs are interactive. Mouse over each data point and additional data will be displayed. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
overall <- ggplot(data = total, aes(x = Group.1, y = x/1000, color = type, label = type)) +
  geom_line(aes(group = type), size=1) + ylab("Dollar Amount (in thousands)") + xlab("Academic Year") + ggtitle("Overall Trends")

cahplot <- ggplot(data = cah, aes(x = Year, y = Amount/1000, color = Dept, label = Dept)) +
  geom_line(aes(group = Dept), size=1) + geom_hline(yintercept = 0, linetype = "longdash") + ylab("Profit/Loss (in thousands)") + ggtitle("College of Arts and Humanities")

cosplot <- ggplot(data = cos, aes(x = Year, y = Amount/1000, color = Dept, label = Dept)) +
  geom_line(aes(group = Dept), size=1) + geom_hline(yintercept = 0, linetype = "longdash") + ylab("Profit/Loss (in thousands)") + ggtitle("College of Science")

cepsplot <- ggplot(data = ceps, aes(x = Year, y = Amount/1000, color = Dept, label = Dept)) +
  geom_line(aes(group = Dept), size=1) + geom_hline(yintercept = 0, linetype = "longdash") + ylab("Profit/Loss (in thousands)")+ ggtitle("College of Education and Professional studies")
busplot <- ggplot(data = bus, aes(x = Year, y = Amount/1000, color = Dept, label = Dept)) +
  geom_line(aes(group = Dept), size=1) + geom_hline(yintercept = 0, linetype = "longdash") + ylab("Profit/Loss (in thousands)")+ ggtitle("College of Business")

```

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12}
ggplotly(overall)
ggplotly(cahplot)
ggplotly(cosplot)
ggplotly(cepsplot)
ggplotly(busplot)
```

### Each Department Individually

These are merely each indivdiual department getting its own graph in order to see more clearly the trend for each department. The horizontal dashed line is the break even point. The y-axis is consistent for each college in order to visually compare the size of each profit or loss.

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = cah, aes(x = Year, y = Amount)) + geom_line(aes(group = Dept)) + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Profit or Loss") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Arts and Humanities")

ggplot(data = cos, aes(x = Year, y = Amount)) + geom_line(aes(group = Dept)) + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Profit or Loss") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Science")

ggplot(data = ceps, aes(x = Year, y = Amount)) + geom_line(aes(group = Dept)) + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Profit or Loss") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Education and Professional studies")

ggplot(data = bus, aes(x = Year, y = Amount)) + geom_line(aes(group = Dept)) + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Profit or Loss") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Business")
```

###Decline in Personnel Expenses by College and Department

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = cah, aes(x = Year, y = Personnel/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Personnel Expenses (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Arts and Humanities")


ggplot(data = cos, aes(x = Year, y = Personnel/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Personnel Expenses (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Science")

ggplot(data = ceps, aes(x = Year, y = Personnel/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Personnel Expenses (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Education and Professional Studies")

ggplot(data = bus, aes(x = Year, y = Personnel/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Personnel Expenses (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Business")
```

###Decline in Net Tuition by College and Department

```{r message=FALSE, warning=FALSE, echo=FALSE}
ggplot(data = cah, aes(x = Year, y = Net.Tuition/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Net Tuition (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Arts and Humanities")


ggplot(data = cos, aes(x = Year, y = Net.Tuition/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Net Tuition (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash")+ ggtitle("College of Science")

ggplot(data = ceps, aes(x = Year, y = Net.Tuition/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Net Tuition (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("College of Education and Professional Studies")

ggplot(data = bus, aes(x = Year, y = Net.Tuition/1000)) + geom_bar(aes(group = Dept),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + facet_wrap(~Dept) + ylab("Net Tuition (in thousands)" ) + geom_hline(yintercept = 0, linetype = "longdash")+ ggtitle("Business")
```


```{r message=FALSE, warning=FALSE, echo=FALSE}

first$famount <- first$Personnel
first$lamount <- last$Personnel
first$persondiff <-  1- (first$lamount/first$famount)
first$persondiff2 <-  (first$lamount/first$famount)
first$netfamount <- first$Net.Tuition
first$netlamount <- last$Net.Tuition
first$tuitiondiff <- 1- (first$netlamount/first$netfamount)
first$tuitiondiff2 <- (first$netlamount/first$netfamount)
first$totaldiff <- first$persondiff2 - first$tuitiondiff2
diff <- select(first, Dept, College, persondiff, tuitiondiff, totaldiff)
```

#Changes From AY2011-2015

The following graphs may be the most important. 

Graph 1- Which departments have been cut the most in regard to personnel from AY2011 to AY2015? Those on the far left have seen the largest reductions (by percentage), and those on the far right have been cut the least over that five year time period.

Graph 2- Which departments have had the largest loss in net tuition? Again, those on the left have had the largest declines from AY2011-AY2015. A few departments on the right have seen positive tuition growth. 

Graph 3- This graph depicts the difference between the percentage of personnel reductions and the percentage decrease in net tuition revenue. Net tuition is calculated by taking total tuition and then deducting both waivers and scholarships. Those departments on the far left of the graph have not seen their personnel cuts keep pace with the decrease that they have experienced in net tuition. Those departments on the far right of the graph have seen total net tuition increases but have not seen a corresponding increase in personnel. The vertical dashed line falls directly on the Communication department, as it's drop in personnel expenses has moved almost completely in lock step with it's overall decrease in net tuition, and therefore has been treated the most "fairly" in the last five years.  

```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=8}
ggplot(diff, aes(x=reorder(Dept, -persondiff), y = persondiff*100)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Personnel Cuts from AY2011-AY2015 (Percentage)") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Personnel Reductions (by Percentage) From AY2011-AY2015")

 ggplot(diff, aes(x=reorder(Dept, -tuitiondiff), y = tuitiondiff*100)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Tuition Loss from AY2011-AY2015 (Percentage)") + geom_segment(aes(x = 25, y = -35, xend = 27.5, yend = -25), arrow = arrow(length = unit(.5, "cm"))) + annotate("text", x = 22.5, y = -40, label = "Depts with Positive Tuition Growth") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Tuition Losses (by Percentage) From AY2011-AY2015 ")

ggplot(diff, aes(x=reorder(Dept, -totaldiff), y = totaldiff*100)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) + annotate("text", x = 12.5, y = 55, label = "Depts Where Personnel Cut Haven't Kept Pace w/Tuition Declines") + annotate("rect", xmin = 0, xmax = 26, ymin = 0, ymax = 50,  alpha = .2) +xlab("Department") + ylab("Difference between Personnel Cuts and Tuition Declines") + geom_vline(xintercept = 26, linetype = "longdash") + ggtitle("Differences Between Personnel Reduction and Tuition Losses From AY2011-AY2015" )
```

###Growth Scenarios 

What would each department look like in different growth scenarios? 

Graph 1- Profit and loss for each department in AY2015-2016. The vertical dashed line is the break even point. 

Graph 2- Profit and loss if tuition rose 15% across the board and expenses were held constant. 

Graph 3- Profit and loss if expenses were at the AY2015-2016 level, but tuition returned to the AY2011-2012 level. 


```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=8}
ggplot(last, aes(x=reorder(Dept, -Amount), y = Amount/1000)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Profit or Loss By Department (in Thousands)") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Profit or Loss in AY2015-2016") + geom_vline(xintercept = 17, linetype = "longdash")
ggplot(last, aes(x=reorder(Dept, -tuition15), y = tuition15/1000)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Personnel Cuts from AY2011-AY2015 (in Thousands)") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Profit or Loss if Net Tuition Rose 15%") + geom_vline(xintercept = 20, linetype = "longdash")

compare <- select(first, Dept, College, Net.Tuition)
compare$Personnel <- last$Personnel
compare$diff <- compare$Net.Tuition - compare$Personnel
ggplot(compare, aes(x=reorder(Dept, -diff), y = diff/1000)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Personnel Cuts from AY2011-AY2015 (in Thousands)") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Profit or Loss if Net Tuition was at AY2011-2012 Levels") + geom_vline(xintercept = 29, linetype = "longdash")
```


###How are the Losses Spread Across Departments?

Graph 1- This graph displays total profits or losses across the five year time frame in the data. Departments above the x-axis had profit for the time period. Departments below the x-axis had total losses for the time period. The y-axis is in thousands of dollars as well (i.e. Sociology generated nearly 8 million dollars in profit)

Graph 2- This graph only looks at the departments who had a net loss in the time period and indicates what percentage of the total loss was due to their balance sheet. The y-axis is percentage of total loss. The total loss for these departments was $24.2 million for the time period. 


```{r message=FALSE, warning=FALSE, echo=FALSE, fig.width=12, fig.height=8}
ggplot(agg5, aes(x=reorder(Group.1, -x), y = x/1000)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Dollar Amount (in Thousands)") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Total Profit/Loss by Department From AY2011-AY2015")
losers <- filter(agg5, x < 0)
losers$percent <- losers$x/sum(losers$x)
ggplot(losers, aes(x=reorder(Group.1, -percent), y = percent*100)) + geom_bar(aes(fill=College),stat="identity") + theme(axis.text.x = element_text(angle = 90)) +xlab("Department") + ylab("Percent of Total Loss") + geom_hline(yintercept = 0, linetype = "longdash") + ggtitle("Which Departments Have Accounted for the Biggest Losses?")
```


###Search the Data Yourself

This allows you to look at the raw data yourself. Just type a query into the search box on the right and you can search for specific departments, specific colleges, years, etc. You can also sort by clicking the up or down arrows in each column. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
datatable <- select(all, Dept, College, Year, Personnel, Other.Expenses, Total, Net.Tuition, Amount)
datatable(datatable, colnames = c('Dept', 'College', 'Year', 'Personnel Budget', 'Other Expenses', 'Total Tuition', 'Net Tuition', 'Total Profit or Loss'))

```
